---
title: "GIS layers"
author: '@rvosa'
date: '2022-05-23'
output: html_document
---

```{r setup, include=FALSE}
# here we configure our environment and load the required dependencies
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(raster)
library(tidyverse)
library(rgdal)
```

# Introduction

GIS data releases such as BioClim (WorldClim) and ENVIREM are very useful for
eco-evolutionary research. However, some of the layers are hard to develop an
intuition for based solely on the definition of what the variable means. Here
we provide worldwide maps of the variables. The main purpose of this is to be
of help in interpreting the results of ENM/SDM analyses or other approaches 
that select variables out of larger sets, for example when including OMI-
transformed point estimates of species optima. We therefore map these data
only at the lowest resolution (10 arc minutes) to get a global perspective 
without too much data.

## BioClim

Fick, S.E. and R.J. Hijmans, 2017. WorldClim 2: new 1km spatial resolution 
climate surfaces for global land areas. 
International Journal of Climatology 37 (12): 4302-4315.
[https://doi.org/10.1002/joc.5086](https://doi.org/10.1002/joc.5086)

Data at: [https://www.worldclim.org/data/bioclim.html](https://www.worldclim.org/data/bioclim.html)

Bioclimatic variables are derived from the monthly temperature and rainfall 
values in order to generate more biologically meaningful variables. These are 
often used in species distribution modeling and related ecological modeling 
techniques. The bioclimatic variables represent annual trends (e.g., mean 
annual temperature, annual precipitation) seasonality (e.g., annual range in 
temperature and precipitation) and extreme or limiting environmental factors 
(e.g., temperature of the coldest and warmest month, and precipitation of the 
wet and dry quarters). A quarter is a period of three months (1/4 of the year).

### BIO1 = Annual Mean Temperature

This tracks the latitude: highest near the equator, lowest near the poles.
Mountains are exceptions: higher in the mountains is similar to further
away from the equator.

```{r bio1info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_1.tif")
```
```{r bio1map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_1.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))

Diurnal = through the 24hr cycle. So for every month the difference between max
temp (during the day) and min temp (at night) is taken. This is highest in 
desert regions, with cold nights and hot days.

```{r bio2info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_2.tif")
```
```{r bio2map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_2.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO3 = Isothermality (BIO2/BIO7) (×100)

This variable is BIO2 divided by BIO7, so, the daily range divided by the annual
range. Because the annual range is lowest in the equatorial tropics (there are 
more or less no seasons), this quotient is highest in the tropics.

```{r bio3info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_3.tif")
```
```{r bio3map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_3.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO4 = Temperature Seasonality (standard deviation ×100)

The degree to which temperature varies by season. This is highest in the polar
regions, where summers can be above 10 degrees while winters are freezing. 

```{r bio4info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_4.tif")
```
```{r bio4map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_4.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO5 = Max Temperature of Warmest Month

This mostly shows that extraordinary heat waves can happen almost anywhere 
except high in the polar regions and high in the mountains.

```{r bio5info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_5.tif")
```
```{r bio5map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_5.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO6 = Min Temperature of Coldest Month

No frost in tropics, but possible in arid subtropics. Severe freezing in polar
regions and in the mountains.

```{r bio6info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_6.tif")
```
```{r bio6map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_6.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO7 = Temperature Annual Range (BIO5-BIO6)

The difference between max temperature during the warmest month (BIO5) and the 
min temperature during the coldest. This difference is the greatest in the polar
regions and smallest near the equator.

```{r bio7info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_7.tif")
```
```{r bio7map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_7.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO8 = Mean Temperature of Wettest Quarter

The wettest quarter is warmest in the tropics.

```{r bio8info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_8.tif")
```
```{r bio8map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_8.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO9 = Mean Temperature of Driest Quarter

The driest quarter is warmest in the tropics.

```{r bio9info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_9.tif")
```
```{r bio9map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_9.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO10 = Mean Temperature of Warmest Quarter

Closely tracks BIO1


```{r bio10info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_10.tif")
```
```{r bio10map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_10.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO11 = Mean Temperature of Coldest Quarter

Closely tracks BIO1

```{r bio11info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_11.tif")
```
```{r bio11map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_11.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO12 = Annual Precipitation

I.e. rain, snow, hail, etc. Precipitation roughly follows bands with the 
equatorial tropics being the wettest, then followed by the driest (desert) zone
in either direction away from the equator, then the less dry subtropical zone
(including mediterranean climates) followed by the wetter temperate and boreal
zones.

```{r bio12info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_12.tif")
```
```{r bio12map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_12.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO13 = Precipitation of Wettest Month

Closely tracks BIO12

```{r bio13info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_13.tif")
```
```{r bio13map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_13.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO14 = Precipitation of Driest Month

The general trend is that high values are even more tightly clustered around the
equator, where precipitation is the most constant year round (and therefore even
during the driest month). Besides that, coastal mountain ranges catch the most
constant precipitation.

```{r bio14info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_14.tif")
```
```{r bio14map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_14.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO15 = Precipitation Seasonality (Coefficient of Variation)

Values are the highest, meaning variation is the greatest across seasons, in the
arid zones flanking the wet, equatorial tropics. Example: the Sahel.

```{r bio15info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_15.tif")
```
```{r bio15map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_15.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO16 = Precipitation of Wettest Quarter

Closely tracks BIO13.

```{r bio16info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_16.tif")
```
```{r bio16map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_16.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO17 = Precipitation of Driest Quarter

Closely tracks BIO14.

```{r bio17info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_17.tif")
```
```{r bio17map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_17.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO18 = Precipitation of Warmest Quarter

Highest in tropics and on the wet side of mountain ranges, e.g. see Himalaya.


```{r bio18info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_18.tif")
```
```{r bio18map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_18.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```

### BIO19 = Precipitation of Coldest Quarter

Closely tracks BIO14.


```{r bio19info}
GDALinfo("../data/bioclim/2.1/10m/wc2.1_10m_bio_19.tif")
```
```{r bio19map}
r <- raster("../data/bioclim/2.1/10m/wc2.1_10m_bio_19.tif", values=T, nrows=1080, ncols=2160, xmn=-180, xmx=180, ymn=-90, ymx=90)
plot(r)
```
