---
title: "GIS layers"
author: '@rvosa'
date: '2023-05-23'
output: html_document
---

```{r setup, include=FALSE}
# here we configure our environment and load the required dependencies
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(raster)
library(tidyverse)
library(rgdal)
```

# Introduction

GIS data releases such as BioClim (WorldClim) and ENVIREM are very useful for
eco-evolutionary research. However, some of the layers are hard to develop an
intuition for based solely on the definition of what the variable means. Here
we provide worldwide maps of the variables. The main purpose of this is to be
of help in interpreting the results of ENM/SDM analyses or other approaches 
that select variables out of larger sets, for example when including OMI-
transformed point estimates of species optima. We therefore map these data
only at the lowest resolution (10 arc minutes) to get a global perspective 
without too much data.

## ENVIREM

Title P.O., Bemmels J.B. 2018. ENVIREM: an expanded set of bioclimatic and 
topographic variables increases flexibility and improves performance of 
ecological niche modeling. Ecography. 41:291–307.
[doi:10.1111/ecog.02880](https://doi.org/10.1111/ecog.02880)

Data at: [https://envirem.github.io/](https://envirem.github.io/)

Environmental variables that are thought to be relevant to species' ecology and 
geographic distribution are essential for applications such as species 
distribution modeling. However, the number of such variables that are available, 
that span multiple time periods, and that can easily be integrated with other 
datasets is very limited.

With the ENVIREM dataset, we provide a number of climatic and topographic 
variables that have been described in the literature, and make them available at 
the same resolutions as are available at WorldClim, and for both current and 
past time periods. 

### annualPET

Annual potential evapotranspiration: a measure of the ability of the atmosphere 
to remove water through evapotranspiration processes, given unlimited moisture.

This is high in regions with high actual or _possible_ evaporation. Not 
necessarily tracking the latitudinal bands precisely: areas with high values 
include the Taklamakan desert, and Southern Iberia, for example. 

This measure expresses the potential, so it is unaffected by actual availability 
or lack of water. Which means there is little difference between, for example, 
the dry and wet parts of Africa.

```{r annualPETinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_annualPET.tif")
```
```{r annualPETmap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_annualPET.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### aridityIndexThornthwaite

Thornthwaite aridity index: Index of the degree of water deficit below water 
need. This measure is strongly influenced by how much water is available. Values
are high in areas that are actually dry (Sahara) but low in areas with a lot of
rain (e.g. Congo basin).

```{r aridityIndexThornthwaiteinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_aridityIndexThornthwaite.tif")
```
```{r aridityIndexThornthwaitemap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_aridityIndexThornthwaite.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### climaticMoistureIndex

A metric of relative wetness and aridity. Values are high in wet areas, such as
the west coasts of continents, especially at higher latitudes, or where 
mountains force precipitation (e.g. Alps, Himalayas), or in the wet, equatorial
tropics.

```{r climaticMoistureIndexinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_climaticMoistureIndex.tif")
```
```{r climaticMoistureIndexmap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_climaticMoistureIndex.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### continentality

Average temp. of warmest month - average temp. of coldest month. Values are 
highest in the interior of large continents at high latitudes. The summers in
much of Canada and in Siberia can be surprisingly warm, while the winters are
very cold. Values are lowest around the equator.

```{r continentalityinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_continentality.tif")
```
```{r continentalitymap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_continentality.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### embergerQ

Emberger's pluviothermic quotient: a metric that was designed to differentiate 
among Mediterranean type climates. At global scale, values are highest in the
very wettest of the wet tropics but are lower nearly everywhere else.

```{r embergerQinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_embergerQ.tif")
```
```{r embergerQmap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_embergerQ.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### growingDegDays0

Sum of mean monthly temperature for months with mean temperature greater than 
0℃ multiplied by number of days. Values are high in warm areas and low in cold
areas. Consequently, this measure tracks latitude, except where high mountains
are involved. For example, values are lower in the Andes than their latitude
would suggest.

```{r growingDegDays0info}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_growingDegDays0.tif")
```
```{r growingDegDays0map}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_growingDegDays0.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### growingDegDays5

sum of mean monthly temperature for months with mean temperature greater than 
5℃ multiplied by number of days. Closely tracks growingDegDays0.

```{r growingDegDays5info}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_growingDegDays5.tif")
```
```{r growingDegDays5map}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_growingDegDays5.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### maxTempColdestMonth

Max. temp. of the coldest month. Closely tracks latitude but with some mountain
effects.

```{r maxTempColdestMonthinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_maxTempColdest.tif")
```
```{r maxTempColdestMonthmap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_maxTempColdest.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### minTempWarmestMonth

Min. temp. of the warmest month. Tracks latitude but with more mountain effects.

```{r minTempWarmestMonthinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_minTempWarmest.tif")
```
```{r minTempWarmestMonthmap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_minTempWarmest.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### monthCountByTemp10

Count of the number of months with mean temp greater than 10℃. Tracks latitude
but with strong mountain effects. Very 'stepped' because values are integers
(0-12) and so areas with very low and very high values could be near each other,
e.g. 0 months with temp >10 high in the Andes next to areas where every month
is >10 in the Amazonian rainforest.

```{r monthCountByTemp10info}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_monthCountByTemp10.tif")
```
```{r monthCountByTemp10map}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_monthCountByTemp10.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### PETColdestQuarter

Mean monthly PET of coldest quarter. PET = potential evapotranspiration, a 
measure of the ability of the atmosphere to remove water through 
evapotranspiration processes, given unlimited moisture. Values are highest
in the semi-arid regions, such as the Sahel, next to the equatorial wet tropics.

```{r PETColdestQuarterinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_PETColdestQuarter.tif")
```
```{r PETColdestQuartermap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_PETColdestQuarter.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### PETDriestQuarter

Mean monthly PET of driest quarter. PET = potential evapotranspiration, a 
measure of the ability of the atmosphere to remove water through 
evapotranspiration processes, given unlimited moisture. Values are highest in
semi-arid zones with a dry season, for example in Mediterranean climates such
as the Med itself, the US westcoast, the Cape, southern Chile and western 
Australia but also beyond that in the Maghreb and the Levant/Middle East.

```{r PETDriestQuarterinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_PETDriestQuarter.tif")
```
```{r PETDriestQuartermap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_PETDriestQuarter.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### PETseasonality

Monthly variability in potential evapotranspiration. Values are highest in
'temperate' climates, not too hot and not too cold.

```{r PETseasonalityinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_PETseasonality.tif")
```
```{r PETseasonalitymap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_PETseasonality.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### PETWarmestQuarter

Mean monthly PET of warmest quarter. Values are highest in (semi)arid
(sub)tropical areas. Especially deserts, but also steppes, prairies.

```{r PETWarmestQuarterinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_PETWarmestQuarter.tif")
```
```{r PETWarmestQuartermap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_PETWarmestQuarter.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### PETWettestQuarter

Mean monthly PET of wettest quarter. Values are abruptly lower in Mediterranean
climate zones compared to both adjacent deserts and adjacent temperate zones.

```{r PETWettestQuarterinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_PETWettestQuarter.tif")
```
```{r PETWettestQuartermap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_PETWettestQuarter.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### thermicityIndex

Compensated thermicity index: sum of mean annual temp., min. temp. of coldest 
month, max. temp. of the coldest month, x 10, with compensations for better 
comparability across the globe. Tracks latitude but with mountain effects.

```{r thermicityIndexinfo}
GDALinfo("../data/envirem/1.0/10m/current_10arcmin_thermicityIndex.tif")
```
```{r thermicityIndexmap}
r <- raster("../data/envirem/1.0/10m/current_10arcmin_thermicityIndex.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```