---
title: "GIS layers"
author: '@rvosa'
date: '2022-05-25'
output: html_document
---

```{r setup, include=FALSE}
# here we configure our environment and load the required dependencies
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(raster)
library(tidyverse)
library(rgdal)
```

# Introduction

GIS data releases such as BioClim (WorldClim) and ENVIREM are very useful for
eco-evolutionary research. However, some of the layers are hard to develop an
intuition for based solely on the definition of what the variable means. Here
we provide worldwide maps of the variables. The main purpose of this is to be
of help in interpreting the results of ENM/SDM analyses or other approaches 
that select variables out of larger sets, for example when including OMI-
transformed point estimates of species optima. We therefore map these data
only at the lowest available resolution to get a global perspective 
without too much data.

## HWSD

The Harmonized World Soil Database is a 30 arc-second raster database with over 
15 000 different soil mapping units that combines existing regional and national 
updates of soil information worldwide (SOTER, ESD, Soil Map of China, WISE) with 
the information contained within the 1:5 000 000 scale FAO-UNESCO Soil Map of 
the World (FAO, 1971-1981).

Data source: https://www.fao.org/soils-portal/soil-survey/soil-maps-and-databases/harmonized-world-soil-database-v12/en/

### Aspect

The compass direction that a topographic slope faces, usually measured in 
degrees from north. Aspect can be generated from continuous elevation surfaces.
This is a small-scale pattern, i.e. there are very few global patterns.

```{r BulkDensityinfo}
GDALinfo("../data/hwsd/2.1/10m/Aspect_10deg.tif")
```
```{r BulkDensitymap}
r <- raster("../data/hwsd/2.1/10m/Aspect_10deg.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### Slope

Slope is the percent change in elevation over a certain distance. This is high
in rugged mountains.

```{r Slopeinfo}
GDALinfo("../data/hwsd/2.1/10m/slope_10deg.tif")
```
```{r Slopemap}
r <- raster("../data/hwsd/2.1/10m/slope_10deg.tif", values=T, nrows=900, ncols=2160, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```
