---
title: "GIS layers"
author: '@rvosa'
date: '2022-05-25'
output: html_document
---

```{r setup, include=FALSE}
# here we configure our environment and load the required dependencies
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(raster)
library(tidyverse)
library(rgdal)
```

# Introduction

GIS data releases such as BioClim (WorldClim) and ENVIREM are very useful for
eco-evolutionary research. However, some of the layers are hard to develop an
intuition for based solely on the definition of what the variable means. Here
we provide worldwide maps of the variables. The main purpose of this is to be
of help in interpreting the results of ENM/SDM analyses or other approaches 
that select variables out of larger sets, for example when including OMI-
transformed point estimates of species optima. We therefore map these data
only at the lowest resolution (10 arc minutes) to get a global perspective 
without too much data.

## Global Soil Dataset for use in Earth System Models (GSDE)

Shangguan, W., Dai, Y., Duan, Q., Liu, B. and Yuan, H., 2014. A Global Soil 
Data Set for Earth System Modeling. Journal of Advances in Modeling Earth 
Systems, 6: 249-263. 
doi:[10.1002/2013MS000293](https://doi.org/10.1002/2013MS000293)

From website:

> We developed a comprehensive, gridded Global Soil Dataset for use in Earth 
> System Models (GSDE) and other applications as well. GSDE provides soil 
> information including soil particle-size distribution, organic carbon, and 
> nutrients, etc. and quality control information in terms of confidence level. 
> GSDE is based on the Soil Map of the World and various regional and national 
> soil databases, including soil attribute data and soil maps. We used a 
> standardized data structure and data processing procedures to harmonize the 
> data collected from various sources. We then used a soil type linkage method 
> (i.e. taxotransfer rules) and the polygon linkage method to derive the spatial 
> distribution of soil properties. To aggregate the attributes of different 
> compositions of a mapping unit, we used three mapping approaches: 
> area-weighting method, the dominant soil type method and the dominant binned 
> soil attribute method. In the released gridded dataset, we used the 
> area-weighting method as it will meet the demands of most applications. The 
> dataset can be also aggregate to a lower resolution. The resolution is 30 
> arc-seconds (about 1 km at the equator). The vertical variation of soil 
> property was captured by eight layers to the depth of 2.3 m (i.e. 0- 0.045, 
> 0.045- 0.091, 0.091- 0.166, 0.166- 0.289, 0.289- 0.493, 0.493- 0.829, 0.829- 
> 1.383 and 1.383- 2.296 m).

Data at: http://globalchange.bnu.edu.cn/research/soilw

### BulkDensity

The soil bulk density (BD) is the mass of soil material divided by the total 
volume of solids and pores. Because of lack of data, the BD is usually estimated 
from the sand, silt, and clay contents in LSMs. The BD is inversely related to 
the porosity of a soil. The BD impacts the soil water retention properties and 
soil hydraulic conductivity.

Apparently, soil is less dense at high latitudes.

```{r BulkDensityinfo}
GDALinfo("../data/soil/5m/BulkDensity.tif")
```
```{r BulkDensitymap}
r <- raster("../data/soil/5m/BulkDensity.tif", values=T, nrows=1800, ncols=4320, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### ClayPercentage

In the GSDE, the areas with high sand content were found in the Africa, Canada, 
Australia, Central Asia, and the Arabian Peninsula, most of which are deserts. 
Usually, areas with high sand content had low clay content, and vice versa. 
High clay content appeared in the South America, Central Africa, India, and 
East Australia.

**Note**: there will be a lot of local structure, e.g. clay content should also
be high locally in river valleys.

```{r ClayPercentageinfo}
GDALinfo("../data/soil/5m/ClayPercentage.tif")
```
```{r ClayPercentagemap}
r <- raster("../data/soil/5m/ClayPercentage.tif", values=T, nrows=1800, ncols=4320, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### OrganicCarbon

The soil organic carbon (SOC) increases the water-holding capacity of soil. The 
SOC is also important to plant growth and is a major factor in the overall 
health of soil. Usually, the SOC is not measured directly in a soil survey but 
is estimated from the soil organic matter. In the GSDE, high SOC was primarily 
found in the high latitudes of the Northern Hemisphere. Low SOC was found in the 
desert areas. 

```{r OrganicCarboninfo}
GDALinfo("../data/soil/5m/OrganicCarbon.tif")
```
```{r OrganicCarbonmap}
r <- raster("../data/soil/5m/OrganicCarbon.tif", values=T, nrows=1800, ncols=4320, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```

### PhCaCL

pH, measured in CaCl2 solution. A low pH means an acidic soil, e.g. in the 
Amazonian and Congo basins, but also at high latitudes in Eurasia and North
America. In both cases, high amounts of rain as well as intense decomposition
processes serve to lower the pH compared to more arid regions, which are more
alkaline.

```{r PhCaCLinfo}
GDALinfo("../data/soil/5m/PhCaCL.tif")
```
```{r PhCaCLmap}
r <- raster("../data/soil/5m/PhCaCL.tif", values=T, nrows=1800, ncols=4320, xmn=-180, xmx=180, ymn=-60, ymx=60)
plot(r)
```